Induced universal graphs for families of small graphs

### 2021-10-19 paper/paper.tex
Paper version 2.0

We write $\calF(k)$ to denote the family of all graphs on $k$ vertices,
and we write $f(k)$ to denote the order (that is, the
number of vertices) of a minimal induced universal graph for $\calF(k)$.
We write $F(k)$ to denote the number of non-isomorphic graphs of order $f(k)$
that are induced universal for $f(k)$.
To give an example, we have $f(3)=5$ and $F(3)=5$; all five of the minimal
induced universal graphs
for $\calF(3)$ are shown in \Cref{fig:univ3a}. \Cref{fig:univ3b} shows the four
graphs in $\calF(3)$ as induced subgraphs of a single induced universal graph.

@@

This paper uses a brute-force approach similar to that of Preen to find minimal
induced universal graphs for families of small graphs.  Our contributions
are (1) an ordering strategy to reduce the number of subgraph isomorphism
calls required by the brute force algorithm; (2) exact values for $f(k)$ $(4 \leq k \leq 6)$
and $F(k)$ $(1 \leq k \leq 5)$; (3) an upper bound of 18 for $f(7)$; (4) a hill-climbing
search algorithm to find small (but not necessarily optimal) induced universal graphs
(5) exact results for families of small trees.

@@

Table~\ref{tab:graphresults} shows, for $0 \leq k \leq 5$, the value of $f(k)$.
The table also shows $F(k)$, the number of minimal induced universal
graphs for the family of all $k$-vertex graphs.  To my knowledge, the value of
$f(5)$ and the values of $F(k)$ have not been published previously.

@@

Figure~\ref{fig:graphs} shows examples of minimal induced universal graphs
for the families of all graphs with three, four and five vertices
respectively.

@@

  % Edge: 3 -- 0
  \draw [] (51.166bp,32.248bp) .. controls (44.492bp,40.742bp) and (36.044bp,51.496bp)  .. (29.344bp,60.024bp);
  \draw [] (71.737bp,82.653bp) .. controls (60.758bp,80.98bp) and (46.814bp,78.855bp)  .. (35.882bp,77.189bp);
  % Edge: 4 -- 1
  \draw [] (104.4bp,96.268bp) .. controls (113.37bp,102.97bp) and (124.85bp,111.55bp)  .. (133.86bp,118.28bp);
  % Edge: 4 -- 3
  \draw [] (82.931bp,68.405bp) .. controls (78.728bp,58.107bp) and (73.39bp,45.028bp)  .. (69.205bp,34.773bp);
  \draw (18.0bp,74.46bp) ellipse (18.0bp and 18.0bp);
  \draw (148.63bp,129.31bp) ellipse (18.0bp and 18.0bp);
  \draw (148.0bp,57.0bp) ellipse (18.0bp and 18.0bp);
  \draw (62.36bp,18.0bp) ellipse (18.0bp and 18.0bp);
  \draw (89.87bp,85.42bp) ellipse (18.0bp and 18.0bp);

@@

\caption{Minimal induced universal graphs for the families of all
graphs with 3, 4, and 5 vertices}

@@

    Suppose that $G$ is an induced universal graph for the family of all graphs
    on $k$ vertices, and that $G$ has no more than $2k + 1$ vertices.  Graph
    $G$ must have $K_k$ and $I_k$ as induced sugraphs.  This clique and
    independent set may overlap by no more than one vertex, so their union must
    contain at least $2k - 1$ vertices.  Therefore it is possible to partition
    the vertices of $G$ into three sets: a clique $S_1$, an independent set
    $S_2$, and a third set $S_3$ containing at most 2 vertices.

@@

Each possible edge that is not involved in either the clique or the independent
set --- shown in the figure with a dot --- is added with probability $1/2$.
The loop in \linerangeref{ImproveLoop}{ImproveLoopEnd} then carries out up to
$\AlgVar{maxIter}$ iterations of a step that randomly modifies graph $G$,
and accepts the modification if it causes the number of graphs in $\calF(6)$ that
are induced subgraphs of $G$ to increase.
The attempted-improvement step chooses a random pair of vertices $\{v, w\}$ that
are not involved in either the clique or the independent set
highlighted in \Cref{fig:heuristic-regions}, and
``flips'' the status of the edge between these two vertices: the edge is added if it
not present, and removed otherwise.
After each flip, we count the number of graphs on 6 vertices (from a total
of 156 graphs) that are
isomorphic to an induced subgraph of our 14-vertex graph.  If the most recent flip
increase this number or left it unchanged, we accept it.  After $\AlgVar{maxIter}$
flips, we restart the algorithm with a new random graph.

Our implementation has two optimisations that are not shown in \Cref{alg:heuristic}.
First, scores are cached in a dictionary data structure, so that we can quickly look
up the score of a graph that has already been visited. This data structure is
cleared on return from $\FuncSty{FindInducedUniversalGraph'}$.
(An alternative approach, which we have not implemented, would be avoid
revisiting graphs using a tabu list \cite{DBLP:books/daglib/0093574}.)

@@

We verified that the graphs shown as adjacency matrices in \Cref{fig:adjmat14}
and \Cref{fig:adjmat18} are induced universal graphs for their corresponding families
using a shell script that calls the Glasgow Subgraph Solver to check for the
inclusion of every graph of $k$ vertices \cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/gg/McCreeshP020}.
This script shares no code with the Python program that generated the graphs.

TODO: verification of small instances from the tables using Glasgow Subgraph Solver

@@

\begin{figure}[htb]
    \centering
\begin{tikzpicture}[>=latex',line join=bevel,scale=.35]
  \pgfsetlinewidth{.5bp}
%%
\pgfsetcolor{black}
  % Edge: 6 -- 0
  \draw [] (33.468bp,122.33bp) .. controls (44.467bp,128.89bp) and (59.197bp,137.66bp)  .. (70.288bp,144.27bp);
  % Edge: 6 -- 1
  \draw [] (30.834bp,100.1bp) .. controls (39.774bp,91.035bp) and (51.646bp,78.993bp)  .. (60.571bp,69.941bp);
  % Edge: 7 -- 0
  \draw [] (108.35bp,211.0bp) .. controls (103.69bp,198.96bp) and (97.434bp,182.81bp)  .. (92.754bp,170.73bp);
  % Edge: 7 -- 2
  \draw [] (128.78bp,216.26bp) .. controls (138.81bp,207.93bp) and (152.29bp,196.74bp)  .. (162.31bp,188.42bp);
  % Edge: 7 -- 3
  \draw [] (111.85bp,245.92bp) .. controls (109.9bp,257.68bp) and (107.35bp,272.94bp)  .. (105.4bp,284.67bp);
  % Edge: 8 -- 0
  \draw [] (134.88bp,109.4bp) .. controls (124.44bp,118.9bp) and (110.07bp,131.98bp)  .. (99.64bp,141.47bp);
  % Edge: 8 -- 1
  \draw [] (132.46bp,88.608bp) .. controls (119.73bp,81.799bp) and (101.96bp,72.297bp)  .. (89.227bp,65.491bp);
  % Edge: 8 -- 2
  \draw [] (154.41bp,114.43bp) .. controls (159.1bp,127.91bp) and (165.58bp,146.51bp)  .. (170.24bp,159.91bp);
  % Edge: 8 -- 4
  \draw [] (166.38bp,93.965bp) .. controls (179.82bp,91.611bp) and (198.11bp,88.409bp)  .. (211.53bp,86.058bp);
  % Edge: 8 -- 5
  \draw [] (152.57bp,79.558bp) .. controls (155.7bp,66.447bp) and (159.96bp,48.609bp)  .. (163.09bp,35.509bp);
  % Node: 0
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (86.17bp,153.73bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 1
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (73.34bp,56.99bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 2
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (176.16bp,176.91bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 3
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (102.42bp,302.6bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 4
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (229.48bp,82.92bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 5
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (167.27bp,18.0bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 6
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (18.0bp,113.12bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 7
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (114.87bp,227.81bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 8
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (148.38bp,97.12bp) ellipse (18.0bp and 18.0bp);
\end{scope}
%
\end{tikzpicture}

>>>>

We write $\calF(k)$ to denote the family of all graphs on $k$ vertices,
and we write $f(k)$ to denote the order (that is, the
number of vertices) of a minimal induced universal graph for $\calF(k)$.
We write $F(k)$ to denote the number of non-isomorphic graphs of order $f(k)$
that are induced universal for $\calF(k)$.
To give an example, we have $f(3)=5$ and $F(3)=5$; all five of the minimal
induced universal graphs
for $\calF(3)$ are shown in \Cref{fig:univ3a}. \Cref{fig:univ3b} shows the four
graphs in $\calF(3)$ as induced subgraphs of a single induced universal graph.

@@

This paper uses a brute-force approach similar to that of Preen to find minimal
induced universal graphs for families of small graphs.  Our contributions
are (1) an ordering strategy to reduce the number of subgraph isomorphism
calls required by the brute force algorithm; (2) exact values for $f(k)$ $(5 \leq k \leq 6)$
and $F(k)$ $(1 \leq k \leq 5)$; (3) an upper bound of 18 for $f(7)$; (4) a hill-climbing
search algorithm to find small (but not necessarily optimal) induced universal graphs
(5) exact results for families of small trees.

@@

Recall that $f(k)$ is the order of a minimal induced universal graph for
$\calF(k)$, and $F(k)$ is the number of non-isomorphic graphs of that order
for $\calF(k)$ that are induced universal for $\calF(k)$.  This section
presents values of $f(k)$ and $F(k)$ for $0 \leq k \leq 5$, computed
using \Cref{alg:brute-force}. The results are shown in 
Table~\ref{tab:graphresults}.  To my knowledge, the value of
$f(5)$ and the values of $F(k)$ have not been published previously.

@@

Figure~\ref{fig:graphs} shows examples of minimal induced universal graphs
for the families of all graphs with four and five vertices
respectively.

@@

@@

\caption{Examples of minimal induced universal graphs for $\calF(4)$ and $\calF(5)$}

@@

    Suppose that $G$ is an induced universal graph for the family of all graphs
    on $k$ vertices, and that $G$ has no more than $2k + 1$ vertices.  Graph
    $G$ must have $K_k$ and $I_k$ as induced sugraphs.  Clearly, this clique and
    independent set may overlap by no more than one vertex, so their union must
    contain at least $2k - 1$ vertices.  Therefore it is possible to partition
    the vertices of $G$ into three sets: a clique $S_1$, an independent set
    $S_2$, and a third set $S_3$ containing at most 2 vertices.

@@

Each possible edge that is not involved in either the clique or the independent
set --- shown in the figure with a dot --- is added with probability $1/2$.
The loop in \linerangeref{ImproveLoop}{ImproveLoopEnd} then carries out up to
$\AlgVar{maxIter}$ iterations of a step that randomly modifies graph $G$,
and accepts the modification if it causes the number of graphs in $\calF(6)$ that
are induced subgraphs of $G$ to increase.
The attempted-improvement step chooses a random pair of vertices $\{v, w\}$,
both of which are not in the same coloured square of 
in \Cref{fig:heuristic-regions}, and
``flips'' the status of the edge between these two vertices: the edge is added if it
not present, and removed otherwise.
After each flip, we count the number of graphs on 6 vertices (from a total
of 156 graphs) that are
isomorphic to an induced subgraph of our 14-vertex graph.  If the most recent flip
increase this number or left it unchanged, we accept it.  After $\AlgVar{maxIter}$
flips, we restart the algorithm with a new random graph.

Our implementation has two optimisations that are not shown in \Cref{alg:heuristic}.
First, scores are cached in a dictionary data structure, so that we can quickly look
up the score of a graph that has already been visited. This data structure is
cleared on return from $\FuncSty{FindInducedUniversalGraph'}$.
(An alternative approach, which we have not implemented, would be to
use a tabu list \cite{DBLP:books/daglib/0093574} to avoid revisiting graphs
that have been visited recently.)

@@

We verified that the graphs shown as adjacency matrices in \Cref{fig:adjmat14}
and \Cref{fig:adjmat18} are induced universal graphs for their corresponding families
using a shell script that calls the Glasgow Subgraph Solver \cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/gg/McCreeshP020} to check for the
inclusion of every graph of $k$ vertices.
This script shares no code with the Python program that generated the graphs.

Finally, we verified all but the final row in each of \Cref{tab:graphresults}
and \Cref{tab:treeresults} using a second shell script that calls the Glasgow
Subgraph Solver.  Again, this shares no code with the program used to generate
the results in the tables.

@@

\begin{figure}[htb]
    \centering
\begin{tikzpicture}[>=latex',line join=bevel,scale=.28]
  \pgfsetlinewidth{.5bp}
%%
\pgfsetcolor{black}
  % Edge: 6 -- 0
  \draw [] (33.468bp,122.33bp) .. controls (44.467bp,128.89bp) and (59.197bp,137.66bp)  .. (70.288bp,144.27bp);
  % Edge: 6 -- 1
  \draw [] (30.834bp,100.1bp) .. controls (39.774bp,91.035bp) and (51.646bp,78.993bp)  .. (60.571bp,69.941bp);
  % Edge: 7 -- 0
  \draw [] (108.35bp,211.0bp) .. controls (103.69bp,198.96bp) and (97.434bp,182.81bp)  .. (92.754bp,170.73bp);
  % Edge: 7 -- 2
  \draw [] (128.78bp,216.26bp) .. controls (138.81bp,207.93bp) and (152.29bp,196.74bp)  .. (162.31bp,188.42bp);
  % Edge: 7 -- 3
  \draw [] (111.85bp,245.92bp) .. controls (109.9bp,257.68bp) and (107.35bp,272.94bp)  .. (105.4bp,284.67bp);
  % Edge: 8 -- 0
  \draw [] (134.88bp,109.4bp) .. controls (124.44bp,118.9bp) and (110.07bp,131.98bp)  .. (99.64bp,141.47bp);
  % Edge: 8 -- 1
  \draw [] (132.46bp,88.608bp) .. controls (119.73bp,81.799bp) and (101.96bp,72.297bp)  .. (89.227bp,65.491bp);
  % Edge: 8 -- 2
  \draw [] (154.41bp,114.43bp) .. controls (159.1bp,127.91bp) and (165.58bp,146.51bp)  .. (170.24bp,159.91bp);
  % Edge: 8 -- 4
  \draw [] (166.38bp,93.965bp) .. controls (179.82bp,91.611bp) and (198.11bp,88.409bp)  .. (211.53bp,86.058bp);
  % Edge: 8 -- 5
  \draw [] (152.57bp,79.558bp) .. controls (155.7bp,66.447bp) and (159.96bp,48.609bp)  .. (163.09bp,35.509bp);
  % Node: 0
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (86.17bp,153.73bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 1
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (73.34bp,56.99bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 2
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (176.16bp,176.91bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 3
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (102.42bp,302.6bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 4
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (229.48bp,82.92bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 5
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (167.27bp,18.0bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 6
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (18.0bp,113.12bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 7
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (114.87bp,227.81bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 8
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (148.38bp,97.12bp) ellipse (18.0bp and 18.0bp);
\end{scope}
%
\end{tikzpicture}
